Gradual Soundness: Lessons from Static Python

### 2021-12-11 pj-7.1/paper.tex
pj: multiple inheritance, sp has no vtable check

Our model of \SP{} includes the key aspects of its types,
statements, and values.
There are two main goals:
illustrate the boundaries where typed and untyped code can mix,
and to show how to soundly protect these boundaries.
Compared to other gradual languages, these boundaries are quite restrictive---which
partly explains \SP{} performance~(\cref{s:eval}).
Indeed the model is nearly a subset of model for Nom~\cite{mt-oopsla-2021},
hence we focus on explaining the gaps and omit formal proofs.

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle soundly (FILL examples ... while loop?);
omitting these lets us focus on the difficult corners.
Other Python features are left out because their \SP{} semantics is identical
to Python.
These include \code{eval}, first-class classes, multiple inheritance, and module-level cells.
\SP{} does not rely on types for these features to guide program transformations;
see \cref{s:impl} for further discussion.

@@

\SP{} does not ascribe types to the following Python features.
These are not covered in our model because the implementation simply
gives them the dynamic type and lets the runtime treat them as untyped
Python code.
For each one, the dynamic type is a reasonable choice; static types are
likely to cause more trouble than helpful structure.

@@

\SP{} does not attempt to type first-class classes:
partly because they see little use at Instagram (FILL how much?)
and partly because it is unclear how to incorporate them into
the nominal type system.
The straightforward but restrictive approach is to force code
that uses a first-class class to expect instances of a particular
static class.
FILL is the bracha cook mixin pattern a way to work within nominal~\cite{bc-oopsla-1990}?
Flatt et al.~\cite{fkf-popl-1998} propose a different method that accommodates mixins,
but the method calls a second layer of interface types.
Muehlboeck and Tate~\cite{mt-oopsla-2021} similarly allow structural objects with interfaces.
Adding these layers to \SP{} is an open question; in particular,
prior work does not handle multiple inheritance.

@@

Python allows multiple inheritance.
Classes can inherit from any number of other classes.
To resolve method calls, Python traverses the class hierarchy
to find the first matching receiver (MRO).

[FILL]
\SP{} lets most programs use multiple inheritance.
The one constraint is that a class cannot have two
values bound to the same member name, to ensure v-table
integrity.

>>>>

Our model of \SP{} includes the key aspects of its types,
statements, and values.
The model has two main goals:
illustrate the boundaries where typed and untyped code can mix,
and show how to soundly protect these boundaries.
Compared to other gradual languages, these boundaries are quite restrictive---which
partly explains \SP{} performance~(\cref{s:eval}).
Indeed, the model is nearly a subset of the model for Nom~\cite{mt-oopsla-2021},
hence we focus on explaining the gaps and omit formal proofs.

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle soundly (FILL examples ... while loop?).
Omitting these lets us focus on the difficult corners.
Other Python features are left out because their \SP{} semantics is identical
to Python.
These include \code{eval}, first-class classes, multiple inheritance, and module-level cells.
\SP{} does not rely on types for these features to guide program transformations;
see \cref{s:impl} for further discussion.

@@

\SP{} does not ascribe types to the following Python features.
These are not covered in our model because the implementation simply
assigns the dynamic type and lets the runtime treat them as untyped
Python code.
For each, we claim that the dynamic type is a reasonable choice;
accurate static types would be difficult to maintain.

@@

\SP{} does not attempt to type first-class classes:
partly because they see little use at Instagram (FILL how much?)
and partly because it is unclear how to incorporate them into
the nominal type system.
The straightforward but restrictive approach is to force code
that uses a first-class class to expect instances of a particular
static class.
FILL is the bracha cook mixin pattern a way to work within nominal~\cite{bc-oopsla-1990}?
Flatt et al.~\cite{fkf-popl-1998} propose a different method that accommodates mixins,
but the method calls a second layer of interface types.
Muehlboeck and Tate~\cite{mt-oopsla-2021} similarly allow structural objects with interfaces.
Adding these layers to \SP{} is an open question; in particular,
Python's multiple inheritance would complicate the mixin story.

@@

Python allows classes to inherit from a list of parents.
When resolving a method call to such a class, Python
dynamically traverses the parent list in a fixed order
seeking a first match.

Due to the dynamic method resolution order (MRO),
\SP{} does not track types for classes with multiple parents.
Method calls to these classes execute the same way as
in Python, with no type-directed optimizations to speed up dispatch.
